---
title: "SDMs of the invasive Kelp (Undaria pinnatifida)"
author: "Nestor M. Robinson"
date: "17 May 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
setwd("C:/Users/robinsonnm/Documents/Robinson_2020")# set location of the working directory
memory.limit(50000)
```
getwd()

```{r, include=FALSE}

## Outline ##

This tutorial illustrates how to build, evaluate and project habitat suitability 
for species distribution models within R. It applies to the Asian Kelp in New Zealand coast, 
however it can be easily replicated for other marine species. 

The main steps described bellow, include the following:

1. Data preparation
 + Study area and manipulating the input data
 + Species occurrence data
   - Presence-only data
   - Presence/Absence data
2. Model fitting, prediction, and evaluation
3. Making projections (current and future climatic scenarios)

# 

# Workshop environment preparation # 


{library(biomod2)
library(sp)
library(rgdal)
library(rasterVis)
library(Cairo)
library(raster)
}
```
```{r, include=FALSE}
# custom values for plotting figures # 
{
myCRS <- ' +proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'
NZbox <- extent(164,186,-54,-33)


print_coast <- latticeExtra::layer(sp.lines(coast_sp, lwd=0.05, col='black',alpha=0.8))

}

myTheme=theme_bw()+
  theme(axis.title.x = element_blank(),
        axis.text.x  = element_text(family="Times",color="Black",vjust=0.5,size=10),
        axis.title.y = element_blank(),
        axis.text.y  = element_text(family="Times",colour="Black",
                                    size=10,
                                    vjust=0.5,
                                    hjust=0),
        panel.grid.major = element_blank(), 
        panel.grid.minor =element_blank(),
        plot.title=element_blank(),
        legend.position = c(0.8,0.2),
        legend.title = element_text(size=10,family="Times"),
        legend.text=element_text(size = 10, family = 'Times'),
        legend.key.size = unit(0.6, 'cm'),
        legend.key.width = unit(0.6, 'cm'),
        legend.background = element_blank())

```
## SDMs Asian kelp distribution in New Zealand ##

# 1. Three occurrence datasets were used to run the models. 
a) Presence - background (PB) full dataset: 9,326 presence points 
b) Presence - background unique sites with data (USWD): 66 presence grid cells + 200 pseudoabsence cells
c) Presence - Absence (PA) : 66 presence grid cells + 112 true absence cells


```{r, include= TRUE}
summary(dat1)

# Reading files # 
dat1 <- read.csv("./data_in/occ_dat/occ_swd_dat.csv", header=T) # presence records
dat2 <- read.csv("./data_in/occ_dat/abs_swd_dat.csv", header=T) # absence records

env2000a <- stack ("./data_in/stacks_1km/env2000a_200m.grd")
env2000b <- stack ("./data_in/stacks_1km/env2000b_200m.grd")
rcp2.6_2050 <-stack('./data_in/stacks_1km/rcp2.6_2050.grd')
rcp4.5_2050 <-stack('./data_in/stacks_1km/rcp4.6_2050.grd')
rcp6.0_2050 <-stack('./data_in/stacks_1km/rcp6.0_2050.grd')
rcp8.5_2050 <-stack('./data_in/stacks_1km/rcp8.5_2050.grd')
rcp2.6_2100 <-stack('./data_in/stacks_1km/rcp2.6_2050.grd')
rcp4.5_2100 <-stack('./data_in/stacks_1km/rcp4.6_2050.grd')
rcp6.0_2100 <-stack('./data_in/stacks_1km/rcp6.0_2050.grd')
rcp8.5_2100 <-stack('./data_in/stacks_1km/rcp8.5_2050.grd')


```


# 5. Adjusting the settings for the three models (MAXENT.Phillips, GBM, RF) 
```{r, include=TRUE}

options=BIOMOD_ModelingOptions(
  
  GBM = list( distribution = 'bernoulli',
              n.trees = 5000,
              interaction.depth = 2,
              n.minobsinnode = 5,
              shrinkage = 0.001,
              bag.fraction = 0.5,
              train.fraction = 1,
              cv.folds = 3,
              keep.data = FALSE,
              verbose = FALSE,
              perf.method = 'cv'),
  
  
  RF = list( do.classif = TRUE,
             ntree = 2500,
             mtry = 'default',
             nodesize = 5,
             maxnodes = NULL),
  
  

  MAXENT.Phillips = list( path_to_maxent.jar = './maxent.jar/',
                          memory_allocated =  1024,
                          background_data_dir = 'default',
                          maximumbackground = 'default',
                          maximumiterations = 500,
                          visible = FALSE,
                          linear = FALSE,
                          quadratic =TRUE,
                          product = TRUE,
                          threshold = FALSE,
                          hinge = FALSE,
                          lq2lqptthreshold = 80,
                          l2lqthreshold =5,
                          hingethreshold = 15,
                          beta_threshold = -1,
                          beta_categorical = -1,
                          beta_lqp = -1,
                          beta_hinge = 0.5,
                          betamultiplier = 1,
                          defaultprevalence = 0.5))

```

# 6. Runing the models 
## 6a Run an initial model with presence only and pseudoabsence (PO)
```{r, include=TRUE}
# Manipulate the presence data to get in Biomod2 format

myRespName1='Asian kelp' 
myResp1=(as.numeric(dat1[,"occ"]))
myRespXY1=dat1[,c("x", "y")]
w <- as.numeric(dat1[, "weight"])
v <- rep (1, times= 1200)
myWeights <- c(w, v)

# Formating the data # 
Data1=BIOMOD_FormatingData(resp.var = myResp1,
                                  expl.var=env2000b,
                                  resp.xy=myRespXY1,
                                  resp.nam= 'Asian kelp',
                                  PA.nb.rep = 1,
                                  PA.strategy = 'sre',
                                  PA.sre.quant = 0.025,
                                  PA.nb.absences = 1200,
                                  na.rm = T)
Data1
plot(Data1)


```


```{r, include=TRUE, eco=FALSE}
# Run the models (RF, GBM, and Maxent.Phillips ) # 

m1=BIOMOD_Modeling(
  Data1,
  models=c('MAXENT.Phillips','GBM','RF'),
  models.options = options,
  NbRunEval=1,
  #prevalence= 0.5,
  VarImport = 2,
  Yweights= myWeights,
  models.eval.meth = c('ROC', 'TSS'),
  SaveObj = TRUE,
  rescal.all.models = T,
  do.full.models = T,
  modeling.id = "PO_2000")

# projection over NZ extent under current conditions
Proj01 <- BIOMOD_Projection(
  modeling.output = m1,
  new.env =env2000a,
  proj.name = 'PO_2000',
  selected.models = 'all',
  binary.meth = 'TSS',
  compress = 'xz',
  build.clamping.mask = T,
  output.format = '.grd')

plot(Proj01, str.grep=c('MAXENT.Phillips', 'GBM', 'RF'))
# Esenmble the projections # 

EM01 <- BIOMOD_EnsembleModeling(
  modeling.output = m1,
  chosen.models = 'all',
  em.by='all',
  eval.metric = c('TSS'),
  eval.metric.quality.threshold = c(0.7),
  prob.mean = F,
  prob.cv = F,
  prob.ci = F,
  prob.ci.alpha = 0.05,
  prob.median = F,
  committee.averaging = F,
  prob.mean.weight = T,
  prob.mean.weight.decay = 'proportional' )

EF01 <- BIOMOD_EnsembleForecasting(projection.output = Proj01,
                                   EM.output = EM01)

plot(EF01)
EF01<-get_predictions(EF01)

writeRaster(EF01, './data_out/preds/PO_proj2000.grd', format='raster', overwrite=T)



# Get model evaluation results# 


myBiomodModelEval <- get_evaluations(m1)
myBiomodModelEval["ROC", "Testing.data",,,]
myBiomodModelEval["TSS", "Testing.data",,,]


# Get the variables importance for each model and model run (1-5)#
varImp=as.data.frame(get_variables_importance(m1))

varImp

# Plot response curves

# Load the models for which we want to extract the predicted response curves

myMaxent1 <- BIOMOD_LoadModels(m1, models='GBM')

# plot response plots

myRespPlot1=response.plot2(model=myMaxent1,
               Data =get_formal_data(m1,'expl.var'),
               show.variables= get_formal_data(m1,'expl.var.names'),
               save.file="yes",
               name="response_curve_Maxent",
               plot=TRUE,
               fixed.var.metric = 'median',
               col = c("blue", "red"),
               legend = TRUE,
               data_species = get_formal_data(m1,'resp.var'))
```


# 6c. Models with unique sites with data and true absence (PA)
```{r, include=TRUE, eco=FALSE}
PAdata=rbind(dat1, dat2)


myRespName2='species' 
myResp2=as.numeric(PAdata[,"occ"])
myRespXY2=PAdata[,c("x", "y")]
myWeights2 <- PAdata[-608, "weight"]

str(myWeights2)
# Formating the data # 

PAdata
Data2 <- BIOMOD_FormatingData(resp.var = myResp2,
                                     expl.var = env2000b,
                                     resp.xy = myRespXY2,
                                     resp.name = myRespName2,
                                     PA.strategy = 'user.defined',
                                     #PA.table = PAdata,
                                     na.rm = T)

Data2
plot(Data2)
# Run the models (RF, GBM, and Maxent.Phillips ) # 

m2=BIOMOD_Modeling(
  Data2,
  models=c('MAXENT.Phillips','GBM','RF'),
  models.options = options,
  NbRunEval=1,
  VarImport = 2,
  Yweights = myWeights2,
  #DataSplit = 75,
  models.eval.meth = c('ROC', 'TSS'),
  SaveObj = TRUE,
  rescal.all.models = T,
  do.full.models = T,
  modeling.id = "AllVars_2000_swd2")


# projection over NZ under current conditions
Proj02<- BIOMOD_Projection(
  modeling.output = m2,
  new.env = env2000a,
  proj.name = 'PA_2000',
  selected.models = 'all',
  binary.meth = 'TSS',
  compress = 'xz',
  build.clamping.mask =T,
  output.format = '.grd',
  silent=T)

# Esenmble the projections # 

EM02 <- BIOMOD_EnsembleModeling(
  modeling.output = m2,
  chosen.models = 'all',
  em.by='all',
  eval.metric = c('TSS'),
  eval.metric.quality.threshold = c(0.7),
  prob.mean = F,
  prob.cv = F,
  prob.ci = F,
  prob.ci.alpha = 0.05,
  prob.median = F,
  committee.averaging = F,
  prob.mean.weight = T,
  prob.mean.weight.decay = 'proportional' )

EF02 <- BIOMOD_EnsembleForecasting(projection.output = Proj02,
                                   EM.output = EM02)

plot(EF02)
EF02<-get_predictions(EF02)

writeRaster(EF02, './data_out/preds/PA_proj2000.grd', format='raster', overwrite=T)


# Get model evaluation results#


Eval2 <- get_evaluations(m2)
Eval2["ROC", "Testing.data",,,]
Eval2["TSS", "Testing.data",,,]


# Get the variables importance for each model and model run (1-5)#
varImp2=as.data.frame(get_variables_importance(m2))
varImp2



# Plot response curves

# Load the models for which we want to extract the predicted response curves
myMaxent2 <- BIOMOD_LoadModels(myBiomodModelOut2, models='MAXENT.Phillips')

# plot response plots
myRespPlot2=response.plot2(model=myMaxent2,
                           Data =get_formal_data(myBiomodModelOut2,'expl.var'),
                           show.variables= get_formal_data(myBiomodModelOut2,'expl.var.names'),
                           save.file="yes",
                           name="response_curve_Maxent_swd2",
                           plot=TRUE,
                           fixed.var.metric = 'mean',
                           col = "blue",
                           legend = TRUE,
                           data_species = get_formal_data(myBiomodModelOut2,'resp.var'))




```



# project into the future RCP 2.6 # 
myBiomodRCP2.6_2050 <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut2,
  new.env = RCP2.6_2050,
  proj.name = 'future',
  selected.models = 'all',
  binary.meth = 'TSS',
  compress = 'xz',
  clamping.mask = T,
  output.format = '.grd')

projRCP2.6_2050 <- get_predictions(myBiomodRCP2.6_2050)
plot5=levelplot(projRCP2.6_2050, 
                par.settings=BuRdTheme, 
                main="Presence-background dataset RCP2.6 2050 (74 presence + 113 absence)")+print_coast
plot5

Cairo(file="./figs/Fig3-6.png",
      type='png',
      units='cm',
      width=18,
      height = 12,
      pointsize = 1,
      bg='white',
      dpi=400)
print(plot5)
dev.off()

RangeSize1 <-BIOMOD_RangeSize(myCurrentProj2,projRCP2.6_2050)
plot(RangeSize1$Diff.By.Pixel)
RangeSize1

RangeSize1

Cairo(file="./figs/Fig3-7.png",
      type='png',
      units='cm',
      width=20,
      height = 12,
      pointsize = 1,
      bg='white',
      dpi=300)
plot(RangeSize1$Diff.By.Pixel)
dev.off()
